Profiles of Wind Direction and Speed in Turbulent Ekman Flow

### 2021-10-27 write1/manuscript.tex
completed section 4.4

In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
and the mean velocity is described by a viscous profile of the form
\begin{equation}
  U^{\alpha_\star+}  = z^+
\end{equation}
where the direction of the velocity points into the exact opposite direction of the wall shear stress $\tau$. 
In absence of roughness elements and for small roughness ($z_0^+<5$), this linear regime is known as
viscous sub-layer \cite{foken:m2002,foken:BM1978}.
%
In fact, this law of the wall has no degree of freedom given the drag, i.e. once $u_\star$ and $\alpha_\star$ are defined. 
%
However, theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
aloft.
A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
profile overestimate the velocity in the buffer layer. We therefore introduce a higher-order compensating term into
the linear profile before blending it into the logarithmic law:
\begin{equation}
  U_\text{inner}^{\alpha_\star}^+ = f_{U,\mathrm{inner}}(z^+) = \frac{ z^+ + \gamma_4 (z^+)^{4} + \gamma_6 (z^+)^{6} } { 1 + \gamma_6 /u_{ref} (z^+ )^6}, 
\end{equation} 
where the fourth- and sixth-order terms in the nominator are designed to fit the profile below $y^+\approx 10$ and
the sixth-order correction in the denominator limits the growth of $u_\text{inner}$ for large $y^+$.
%
These correction terms are small for $z^+<5$ where the linear law of the wall
(shown in the inset in Fig.~\ref{fig:u_profiles}) yields an excellent fit of the data; we use here
\begin{equation*}
    \gamma_4 = -3.825 \times 10^{-4};\quad
    \gamma_6 &=&  6.320 \times 10^{-6};\quad 
    u_\mathrm{ref} &=& 1.278 \times 10^{1}.
\end{equation*}

@@

If the difference across the boundary layer is constant ($C_{V0}$) across variations of $\RE$,
the averaged gradient $\partial_{z^+} V^{\alpha_\star+}$ of the spanwise velocity component
must decrease as $1/\delta^+$ as $Re_\tau$ increases.
%
Hence, it should--at a fixed height--be $V^{\alpha_\star} \propto \left(\delta^{+}\right)^{-1}$.
%
A profile that agrees with the constraints of the profile at the surface
and exploits the dependence of $V^{\alpha_\star}$ on $\delta^+$ is 
\begin{equation}
  V^{\alpha_\star}\frac{\delta^+}{G} = f_{V,\mathrm{inner}}(z^+) = v_\mathrm{ref} \left( \omega_v z^+ -1 + \exp[-\omega_v z^+]\right),
  \label{eqn:scaling_spanwise} 
\end{equation}
where $v_\mathrm{ref}$ controls the magnitude of the profile and $\omega_v$ sets the height at which the profile
transitions into an approximately linear one.
%
We find excellent agreement with the DNS data for $500\le\RE_D\le1600$ below $z^+\approx 15$ with
\[v_\mathrm{ref} = 18.85;  \quad \omega_v=0.2353\]
(cf. Fig.~\ref{fig:inner_w}b). 
%
\par
%
Eq.~(\ref{eqn:scaling_spanwise}) establishes a universal mixed scaling for the spanwise velocity:
%
While it requires the vertical coordinate to be expressed in inner units,
the velocity itself is normalized by the geostrophic wind, and becomes
inversely proportional to the friction Reynolds number $\RE_\tau=\delta^+$ when considered at a fixed height.
%
In vicinity of the surface, such mixed scaling has already been identified for higher-order statistics in convective 
flows \citep{mellado:BM2016,li:JAS2018}, where large scales leave their signatures in vicinity of the surface.
%
It is important to note here that, while $V$ is a first-order statistic from a statistical perspective,
the spanwise velocity is a higher-order correction term from the view point of similarity theory and
the channel-flow analogy that is routinely employed in the surface layer.
%
Further, this is consistent with the scaling for the velocity hodograph found in Eq.~(\ref{eqn:hodograph_scaling}) where
the friction velocity also drops out.
%
\par
%
The universal scaling of $V^{\alpha_\star}/G\delta^+$ continues through the surface layer,
and we find a very similar picture to the spanwise component with an increasing
depth of a quasi-logarithmic layer as shown in Fig.~\ref{fig:inner_w}b.
%
We model this layer by a profile of the form
%
\begin{equation}
  f_{V,\mathrm{log}}(z^+) = \frac{V_\mathrm{log}(z^+)}{G} \delta^+ = a_\mathrm{log} + b_\mathrm{log} \log z^+ + c_\mathrm{log} z^+
\end{equation} 
%
This profile is constrained by
\begin{align}
  f_{V,\mathrm{log}}(z^+=10) &=& f_{V,\mathrm{inner}}(z^+=10) &=:& v_{10} \simeq 27.3\\
  \frac{\p}{\p z^+} \left[ f_{V,\mathrm{log}}\right]_{z^+=10} &=& \frac{\p}{\p z^+}\left[ F_{V,\mathrm{inner}}\right]_{z^+=10} &=:&d_{10}\simeq 4.01 \\
  f_{V,\mathrm{log}}(z^+=0.3\delta^+) &=& V^{\alpha_\star}_\mathrm{ek}(z^-=0.3)\delta^+&=:& v_{03} 
\end{align}
where $v_{03}$ is determined by $V_\mathrm{ek}(0.3)$ and $U_\mathrm{ek}(0.3)$ and depends on the Reynolds number.

%
\begin{figure}
  \begin{flushleft}
    \textbf{\ \ a}\hspace{0.45\textwidth}\textbf{\ \ b}\\
  \end{flushleft} 
  \includegraphics[width=1.0\textwidth]{../plot/w_profile.pdf}
  \caption{Profiles of shear-aligned span-wise velocity $(W^\alpha)^+$ versus inner height. 
    Dashed lines show DNS data, thick, opaque lines are from the semi-emprical theory developed above.
    Left panel show standard inner normalization. Right panel shows the inviscid normalization yielding
    a universal profile for the spanwise component of velocity in the inner layer. }
  \label{fig:inner_w}
\end{figure}
\end{subequations}

@@

>>>>

In immediate vicinity to the surface, local turbulent mixing cannot occur for the no-slip/no-penetration boundary condition,
and the mean velocity is described by a viscous profile of the form
\begin{equation}
  U^{\alpha_\star+}  = z^+
\end{equation}
where the direction of the velocity points into the exact opposite direction of the wall shear stress $\tau$. 
In absence of roughness elements and for small roughness ($z_0^+<5$), this linear regime is known as
viscous sub-layer \cite{foken:m2002,foken:BM1978}.
%
In fact, this law of the wall has no degree of freedom given the drag, i.e. once $u_\star$ and $\alpha_\star$ are defined. 
%
However, theoretical foundation is lacking for the exact shape of the velocity profile in the buffer layer; though crucial for turbulence
production, it is commonly understood as a transition region between the linear profile at the surface and the logarithmic profile
aloft.
A pure blending from the linear velocity profile into the logarithmic one is, however, not reasonable as both the linear and logarithmic
profile overestimate the velocity in the buffer layer. We therefore introduce a higher-order compensating term into
the linear profile before blending it into the logarithmic law:
\begin{equation}
  U_\text{inner}^{\alpha_\star}^+ = f_{U,\mathrm{visc}}(z^+) = \frac{ z^+ + \gamma_4 (z^+)^{4} + \gamma_6 (z^+)^{6} } { 1 + \gamma_6 /u_{ref} (z^+ )^6}, 
\end{equation} 
where the fourth- and sixth-order terms in the nominator are designed to fit the profile below $y^+\approx 10$ and
the sixth-order correction in the denominator limits the growth of $u_\text{inner}$ for large $y^+$.
%
These correction terms are small for $z^+<5$ where the linear law of the wall
(shown in the inset in Fig.~\ref{fig:u_profiles}) yields an excellent fit of the data; we use here
\begin{equation*}
    \gamma_4 = -3.825 \times 10^{-4};\quad
    \gamma_6 &=&  6.320 \times 10^{-6};\quad 
    u_\mathrm{ref} &=& 1.278 \times 10^{1}.
\end{equation*}

@@

If the difference across the boundary layer is constant ($C_{V0}$) vs. $\RE$,
the averaged gradient $\partial_{z^+} V^{\alpha_\star+}$ of the spanwise velocity component
must decrease as $1/\delta^+$ with increasing $Re_\tau$. 
Hence, it should--at a fixed height--be $V^{\alpha_\star} \propto \left(\delta^{+}\right)^{-1}$.
A profile that agrees with the constraints of the profile at the surface
and exploits the dependence of $V^{\alpha_\star}$ on $\delta^+$ is 
  V^{\alpha_\star}\frac{\delta^+}{G} = f_{V,\mathrm{visc}}(z^+) = v_\mathrm{ref} \left( \omega_v z^+ -1 + \exp[-\omega_v z^+]\right),
  \label{eqn:scaling_spanwise} 
where $v_\mathrm{ref}$ controls the magnitude of the profile and $\omega_v$ sets the height at which the profile
transitions into an approximately linear one.
We find excellent agreement with the DNS data for $500\le\RE_D\le1600$ below $z^+\approx 15$ with
\[v_\mathrm{ref} = 18.85;  \quad \omega_v=0.2353\]
(cf. Fig.~\ref{fig:inner_w}b). 
    \textbf{\ \ a}\hspace{0.45\textwidth}\textbf{\ \ b}\\
  \includegraphics[width=1.0\textwidth]{../plot/w_profile.pdf}
  \caption{Profiles of shear-aligned span-wise velocity $(W^\alpha)^+$ versus inner height. 
    Dashed lines show DNS data, thick, opaque lines are from the semi-emprical theory developed above.
    Left panel show standard inner normalization. Right panel shows the inviscid normalization yielding
    a universal profile for the spanwise component of velocity in the inner layer. }
  \label{fig:inner_w}
For the adjacent surface layer, we find a log-like transition from the quasi-linear profile
inner profile around $z^+=10$ to a linear profile with increasing $\RE$ (Fig.~\ref{fig:inner_w}b).
We model this transition by 
  f_{V,\mathrm{log}}(z^+) = \frac{V_\mathrm{log}(z^+)}{G} \delta^+ = a_\mathrm{log} + b_\mathrm{log} \log z^+ + c_\mathrm{log} z^+. 
This surface-layer profile matches the inner (viscous) scaling in vicinity of the surface to the outer
(Ekman) scaling above $z^-=0.3$ when constrained by the viscous profile at the bottom and the
Ekman profile at the top:
  f_{V,\mathrm{log}}(z^+=10) &=& f_{V,\mathrm{visc}}(z^+=10) &=:& v_{10} \simeq 27.3\\
  \frac{\p}{\p z^+} \left[ f_{V,\mathrm{log}}\right]_{z^+=10} &=& \frac{\p}{\p z^+}\left[ F_{V,\mathrm{visc}}\right]_{z^+=10} &=:&d_{10}\simeq 4.01 \\
  f_{V,\mathrm{log}}(z^+=0.3\delta^+) &=& V^{\alpha_\star}_\mathrm{ek}(z^-=0.3)\delta^+&=:& v_{03} 
where $v_{03}$ is determined by $V_\mathrm{ek}(0.3)$ and $U_\mathrm{ek}(0.3)$ and depends on $\RE$.
While the $\RE$-dependency of $a_\mathrm{log}$, $b_\mathrm{log}$, $c_\mathrm{log}$
is small, it shows up in Fig.~\ref{fig:profiles} where the normalized profiles
of spanwise velocity become more convex with increasing $\RE$.
We can now quantify this effect by means of the change of $c_\mathrm{log}$ versus $\RE$ which is shown in
Fig.~\ref{fig:clog} (cf. Appendix~\ref{app:log-coefficients};
$a_\mathrm{log}$ and $b_\mathrm{log}$ are then determined by the universal
values of $v_{10}$ and $d_{10}$).
  \includegraphics[width=0.5\textwidth]{../plot/c_log.pdf}
  \caption{c-log.
    \label{fig:clog}}
\subsubsection{Discussion}
Eq.~(\ref{eqn:scaling_spanwise}) establishes a universal mixed scaling for the spanwise velocity
in the viscous layer:
While it requires the vertical coordinate to be expressed in inner units,
the velocity itself is normalized by the geostrophic wind, and becomes
inversely proportional to the friction Reynolds number $\RE_\tau=\delta^+$
when considered at a fixed height.
In vicinity of the surface, such mixed scaling has already been identified for higher-order statistics in convective 
flows \citep{mellado:BM2016,li:JAS2018}, where large scales leave their signatures in vicinity of the surface.
It is important to note here that, while $V$ is a first-order statistic from a statistical perspective,
the spanwise velocity is a higher-order correction term from the perspective of similarity theory and
from the viewpoint of the channel-flow analogy that is routinely employed in the surface layer.
Further, this is consistent with the scaling for the velocity hodograph found in
Eq.~(\ref{eqn:hodograph_scaling}) where the friction velocity also drops out.
In the surface layer, there is not only a mixed scaling--as we had already identified
in the viscous layer--, but we cannot find a universal function onto which the profiles
of spanwise velocity collapse. 
This additional degree of freedom reflects the inner--outer matching problem for the
spanwise velocity, and rather than giving a universal profile for this region,
as is usually done, we resort here to a parametric description of the problem,
namely in terms of the function $f_{V,\mathrm{log}}$ determined by the
parameters $a_\mathrm{log}$, $b_\mathrm{log}$, $c_\mathrm{log}$ which can be estimated
based on the above scaling considerations for any Reynolds number. 
We note that, once the parameter $a_\mathrm{log}$ is known, the parameters $b_\mathrm{log}$
and $c_\mathrm{log}$ can be estimated solely based on $f_{V,\mathrm{visc}}$, i.e. using the
value $v_{10}$ and $d_{10}$ found for the viscous region of the flow.
For the range of Reynolds number relevant to geophysical problems ($10^4\lesssim \RE_D \lesssim 10^6$),
the variation of $c_\mathrm{log}$ is, however, rather small.

@@

\section{Matching the spanwise velocity profiles in the inner layer}
\label{app:log-coefficients}
The spanwise profile in vicinity of the surface is given by
$V/G=f_{V,\mathrm{visc}}\delta ^+$ with
\begin{subequations}
\begin{eqnarray}
  f_{V,\mathrm{visc}} &=& v_\mathrm{ref}\left(\omega_v z^+ -1 + e^{-\omega_v z^+}\right)\\
  f_{V,\mathrm{log}}   &=& a_\mathrm{log} + b_\mathrm{log} \log z^+ + c_\mathrm{log} z^+ 
\end{eqnarray}
\end{subequations}
Matching the profiles and gradient $z_0=10^+$ and the value at $z_1=0.3\dlta^+$ yields
\begin{subequations} 
\begin{eqnarray}
  v_\mathrm{ref} \left(\omega_v z_{0} + e^{-\omega_v z_{0}}\right)
  & = v_{0}=& a_\mathrm{log} + b_\mathrm{log} \log z_{0} +c_\mathrm{log} z_{0}  \\
  &   v_{1}=& a_\mathrm{log} + b_\mathrm{log} \log z_{1} +c_\mathrm{log} z_{1} \\
  v_\mathrm{ref}\omega_z \left( 1 - e^{-\omega_z z_{0}}\right)
  & = d_{0}=& \frac{b_\mathrm{log}}{z_{10}} + c_\mathrm{log} 
\end{eqnarray}
\end{subequations} 
The gradient condition implies $b_\mathrm{log}=(d_{0} -c_\mathrm{log})z_{0}$, and yields
\begin{subequations}
\begin{eqnarray}
  %v_{0}&=& a_\mathrm{log} + z_{0}\left[ (d_0-c_\mathrm{log}) \log z_{0} +c_\mathrm{log}\right]\\
  v_0-z_0d_0\log z_0&=& a_\mathrm{log} + c_\mathrm{log} (z_0- z_0\log z_0) \\
  %v_1 &=& a_\mathrm{log} + z_0(d_0-c_\mathrm{log})\log z_1 +  c_\mathrm{log}z_1\\ 
  v_1-z_0d_0\log z_1 &=& a_\mathrm{log} +c_\mathrm{log} (z_1- z_0\log z_0)  \\
  \Rightarrow c_\mathrm{log}&=& \frac{ \Delta v - z_0d_0 \log z_1/z_0}{\Delta z} 
\end{eqnarray}
with $\Delta z=z_1-z_0$ and $\Delta v=v_1-v_0$.
%
Then, the coefficient $a_\mathrm{log}$ is estimated as 
\begin{eqnarray} 
  a_\mathrm{log} &=& v_0-z_0d_0\log z_0 -  \frac{ \Delta v - z_0d_0 \log z_1/z_0}{\Delta z}\left[
    z_0-z_0\log z_0 
    \right]. 
\end{eqnarray}
We note that $ \log(z_1/z_0) / (z_1-z_0) \rightarrow 0$ for large $z_1$, and as $z_1=0.3\delta^+$,
this implies that the second term in $c_\mathrm{log}$ only plays a role at low and intermediate $\RE$.
Then, $a_\mathrm{log}$ can be estimated as 
\begin{eqnarray}
  a_\mathrm{log} \simeq v_0 -z_0 \left[ d_0 \log  z_0 - \frac{\Delta v}{\Delta z} ( 1- \log z_0) \right] 
\end{eqnarray}
for large $\RE$. 
\end{subequations}
